(Deep) Induction Rules for GADTs\vspace*{-0.1in}

### 2021-08-26 induction-paper/revised-induction-paper.tex
revisions 8-26

\def\lastname{Johann, Ghiorzi, and Jeffries}

@@

\begin{abstract} 
  Deep data types are those that are defined in terms of other such
  data types, including, possibly, themselves.  In that case, they are
  said to be truly nested.  Deep induction is an extension of
  structural induction that traverses {\em all} of the structure in a
  deep data type, propagating predicates on its primitive data
  throughout the entire structure.  Deep induction can be used to prove
  properties of nested types, including truly nested types, that
  cannot be proved via structural induction.  In this paper we show
  how to extend deep induction to deep GADTs that are not truly
  nested. We also show that deep induction cannot be extended to truly
  nested GADTs.\looseness=-1
\end{abstract}

@@

The inductive data types handled by Coq include (possibly mutually
inductive) polynomial algebraic data types (ADTs), and the induction
rules Coq generates for them are the expected ones for standard
structural induction. It has long been understood, however, that these
rules are too weak to be genuinely useful for deep ADTs~\cite{jp20},
i.e., ADTs that are (possibly mutually inductively) defined in terms
of (other) such ADTs.\footnote{A data type is {\em deep} if it is
  (possibly mutually inductively) defined in terms of other such data
  types (including, possibly, itself).}  The following data type of
rose trees, here coded in Agda and defined in terms of the standard
type $\mathsf{List}$ of lists (see
Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:

@@

\noindent
%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
%element of its argument list, and similarly for the lifting
%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
list (resp., rose tree) argument.\footnote{Predicate liftings such as
  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
  supplied as primitives or generated automatically from their
  associated data type definitions as described in
  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
  a container type like $\mathsf{List\,t}$ or $\mathsf{Rose\,t}$
  simply traverses containers of that type and applies its predicate
  argument pointwise to the constituent data of type $\mathsf{t}$.}
Deep induction was also shown in~\cite{jp20} to
% be the missing piece making it possible to
%be the theretofore missing component needed to
% be the essential ingredient making it possible to
deliver the first-ever induction rules --- structural or otherwise ---
for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
types. Deep induction for ADTs and nested types is reviewed in
Section~\ref{sec:ADTs-and-nesteds} below.

This paper shows how to extend deep induction to proper GADTs, i.e.,
to GADTs that are not simply nested types, and thus are not ADTs.  A
constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
nested type, take as arguments data whose types involve instances of
$\mathsf{G}$ other than the one being defined --- including instances
that involve $\mathsf{G}$ itself. But if $\mathsf{G}$ is a proper GADT
then at least one of its constructors will also have such a structured
instance of $\mathsf{G}$ --- albeit one not involving $\mathsf{G}$
itself --- as its codomain.
%But while the types of the arguments to a GADT's constructor can
%include instances of $\mathsf{G}$ that themselves involve
%$\mathsf{G}$, its return type cannot.
For example, the constructor $\mathsf{pair}$ for the GADT

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of pairs, rather than sequences of arbitrary
type, as does $\mathsf{const}$. If one or more of the constructors for
a GADT $\mathsf{G}$ return structured instances of $\mathsf{G}$, then
the GADT will have two distinct, but equally natural, semantics: a
functorial semantics interpreting it as a left Kan
extension~\cite{mac71}, and a parametric semantics interpreting it as
the interpretation of its Church encoding~\cite{atk12,vw10}. As
detailed in~\cite{jgj21}, a key difference in the two semantics is
that the former views GADTs as their {\em functorial
  completions}~\cite{jp19}, and thus as containing more data than just
those expressible in syntax. By contrast, the latter views them as
what might be called {\em syntax-only} GADTs. Happily, these two views
of GADTs coincide for those GADTs that are ADTs or other nested
types. However, both they and their attendant properties differ
greatly for proper GADTs. In fact, the functorial and parametric
semantics for proper GADTs are sufficiently disparate that, by
contrast with the semantics customarily given for ADTs and other
nested types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
to define a functorial parametric semantics for them~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so we quickly realize that GADTs viewed as their
functorial completions cannot possibly support induction rules.  This
makes sense intuitively: induction is a syntactic proof technique, so
of course it cannot be used to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs.  Indeed, this
paper gives the first-ever induction rules --- deep or otherwise ---
for proper GADTs. But it actually delivers far more: it gives a
general framework for deriving deep induction rules for deep GADTs
directly from their syntax. This framework can serve as a basis for
extending modern proof assistants' automatic generation of structural
induction rules for ADTs to automatic generation of deep induction
rules for GADTs. In addition, as for ADTs and other nested types, the
structural induction rule for any GADT can be recovered from its deep
induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

Significantly, deep induction rules for GADTs cannot be derived by
somehow extending the approach of~\cite{jp20} to syntax-only
GADTs. Indeed, the approach taken there makes crucial use of the
functoriality of data types' interpretations from~\cite{jp19}, and
functoriality is precisely what interpreting GADTs as the
interpretations of their Church encodings fails to deliver. Our
approach is to instead first give a predicate lifting styled after
those of~\cite{jp20}, together with a (deep) induction rule, for the
simplest --- and arguably most important --- GADT, namely the equality
GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
the equality GADT to represent $\mathsf{G}$ as its so-called {\em
  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
  ii}) using the predicate liftings for the equality GADT, and any
other GADTs appearing in the definition of $\mathsf{G}$, to
appropriately thread the custom predicates for the primitive types
appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
two-step process delivers deep induction rules for a very general
class of deep GADTs. In Section~\ref{sec:GADTs} we introduce a series
of increasingly complex GADTs as running examples, and in
Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
each of them. In particular, we derive the deep induction rule for the
$\mathsf{Seq}$ data type in~\eqref{eq:seq} in
Section~\ref{sec:ind-seq}. We present our general framework for
deriving (deep) induction rules for (deep) GADTs in
Section~\ref{sec:framework}, and observe that the derivations in
Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
Section~\ref{sec:GADT-nested} we show that, by contrast with truly
nested types, which do have a functorial semantics, syntax-only GADTs'
lack of functoriality means that it is not possible to extend
induction --- deep or otherwise --- to truly nested GADTs. This does
not appear to be much of a restriction, however, since GADTs defined
over themselves do not, to our knowledge, appear in practice or in the
literature. Section~\ref{sec:app} comprises a case study in using deep
induction. All of the deep induction rules appearing in this paper
have been derived by instantiating our general framework. Our Agda
code implementing them is available at~\cite{web-page}.

%\subsection{Related Work}

@@

{\bf Additional Related Work\/} Various techniques for deriving
induction rules for data types that go beyond ADTs have been
studied. For example, Fu and Selinger~\cite{fs18} show, via examples,
how to derive induction rules for arbitrary nested
types. Unfortunately, however, their technique is rather {\em ad hoc},
so is unclear how to generalize it to nested types other than the
specific ones studied there. Moreover,~\cite{fs18} actually derives
induction rules for data types {\em related} to the original nested
types rather than for the original nested types themselves, and it is
unclear whether or not the derived rules are sufficiently expressive
to prove all results about the original nested types that we would
expect to be provable by induction. This latter point echoes the issue
with Coq-derived induction rule for rose trees raised in
Section~\ref{sec:intro}, which has the unfortunate effect of forcing
users to manually write induction (and other) rules for such types for
use in that system. Tassi~\cite{tas19} derives induction rules for
data type definitions in Coq using unary parametricity. His technique
seems to be essentially equivalent to that of~\cite{jp19} for nested
types, although he does not permit true nesting. More recently,
Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
induction rules for nested types. This plugin is also based on unary
parametricity and true nesting still is not permitted.  As far as we
know, no attempts have been made to extend either implementation to
truly nested types or proper GADTs or their deep variants. In fact, we
know of no work other than that reported here that specifically
addresses induction rules for such data types.

@@

It is not possible to extend structural induction to {\em truly}
nested types, i.e., to nested types whose recursive occurrences appear
below themselves. The quintessential example of such a type is that of
bushes\footnote{To define truly nested types in Agda we must use the
  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define functions over
  them we must use the $\mathsf{TERMINATING}$ flag.  (Similar flags
  are required in Coq.) Although as programmers we know from the
  metatheory in~\cite{jp19} that $\mathsf{Bush}$ is well-defined and
  the functions we define over them terminate, the flags are necessary
  because Agda fails to infer these facts. Analogous comments apply at
  several places below.}\cite{bm98}:

@@

\noindent
Even defining a structural induction rule for bushes requires that we
be able to lift the rule's polymorphic predicate argument to
$\mathsf{Bush}$ itself. The more general observation that an induction
rule for any truly nested type must therefore necessarily be a deep
induction rule was, in fact, the original motivation for the
development of deep induction in~\cite{jp20}. The deep induction rule
for bushes is

@@

Although a truly nested type admits only a single induction rule, it
is worth noting that for those nested types that do admit distinct
structural induction and deep induction rules, the latter always
generalize the former. Indeed, the structural induction rule for each
such nested type is recoverable from its deep induction rule by taking
the custom predicates on its data of primitive types to be constantly
$\mathsf{\top}$-valued predicates. This instantiation ensures that the
resulting induction rule only inspects the top-level structure of its
argument, rather than the contents of that structure, which is exactly
what structural induction should do.
%A concrete example of such a derivation is given in
%Section~\ref{sec:ind-equal}.

Under some circumstances deep induction can be mimicked by
hand-threading applications of structural induction through the layers
of data comprising a deep data type. But this is not the case when,
e.g., one or more of the custom predicates in the data type's deep induction
rule cannot be defined by structural induction.

@@

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, each of $\mathsf{pleaf}$ and
$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
regardless of the instances of $\mathsf{PTree}$ appearing in the types
of its arguments. GADTs relax this restriction, allowing their data
constructors both to take as arguments \emph{and return as results}
data whose types involve instances other than the one being
defined. That is, GADTs' constructors' return type instances can, like
that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.\looseness=-1
%these instances can be structured.
%
%Applications of GADTs include generic programming, modeling
%programming languages via higher-order abstract syntax, maintaining
%invariants in data structures, and expressing constraints in embedded
%domain-specific languages. GADTs have also been used, e.g., to
%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

@@

Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
following GADT $\mathsf{LTerm}$, inspired by~\cite{cis194}, is.  It
encodes terms of a simply typed lambda calculus. More robust
variations on $\mathsf{LTerm}$ are, of course, possible, but this
variation is rich enough to illustrate all essential aspects of deep
GADTs --- and later, in Section~\ref{sec:ind-lam}, their deep
induction rules --- while still being small enough to ensure clarity
of exposition.

@@

The equality constraints engendered by GADTs' data constructors makes
deriving (deep) induction rules for them more involved than for ADTs
and other nested types. Nevertheless, we show in this section how to
do so. We first illustrate the key components of our approach by
deriving deep induction rules for the three specific GADTs introduced
in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
abstract these to a general framework that can be applied to any deep
GADT that is not truly nested. As hinted above, the predicate lifting
for the equality GADT plays a central role in deriving both structural
and deep induction rules for more general GADTs.

@@

We can generalize the approach taken in
Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
deep induction rules for deep GADTs. We will treat deep GADTs of the
form

@@

\noindent
for $\mathsf{H^{\wedge}}$. A concrete way to define $\mathsf{H^\wedge
  Map}$ is by induction on the structure of the type $\mathsf{H}$, but
we omit such details since they are not essential to the present
discussion. A further requirement that applies to all of the type
constructors appearing in the right-hand side of the above grammar,
including those in $\ol{\mathsf{K}}$, is that they must all admit
predicate liftings. This is not an overly restrictive condition,
though: all types constructed from sums, products, arrow types and
type application admit predicate liftings, and so do GADTs constructed
from the grammar. In fact, we have seen such liftings for products and
type application in Section~\ref{sec:deep-ind-GADTs}. A concrete way
to define more general predicate liftings is, again, by induction on
the structure of the types. This ensures that the liftings satisfy the
crucial property needed to derive deep induction rules, namely that of
distributing over the type constructors. We do not give a general
definition of predicate liftings here, though, since that would
require us to first design a full type calculus, which is beyond the
scope of the present paper.

@@

In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
derived deep induction rules for GADTs that are not truly
nested. Since both nested types and GADTs without true nesting admit
deep induction rules, we might expect truly nested GADTs to admit them
as well. Unfortunately, however, the techniques developed in the
previous sections do not extend to GADTs. Indeed, the induction rule
for a data type generally relies on (unary) parametricity of the model
interpreting it, and deep induction for a truly nested type or a truly
nested GADT also relies on this interpretation being functorial.
Whereas ADTs and nested types both admit functorial parametric
semantics, proper GADTs admit parametric semantics but do not admit
functorial semantics. In this section we show how the techniques
developed in this paper for deriving deep induction rules go wrong for
truly nested GADTs by analyzing the following very simple
example:\looseness=-1

@@

\begin{equation*}\label{gadt-nested}
\begin{array}{l}
\mathsf{data\ G : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;\;c :\,  \forall \{A : Set\} \to G\,(G\,A) \to G\,(A \times A)}
\end{array}
\end{equation*}
We acknowledge that $\mathsf{G}$ is semantically equivalent to the
empty data type, and thus has a trivial (deep) induction principle. We
could, of course, consider a more realistic counterexample, but this
would only add notational overhead for no gain in conceptual
clarity. Indeed, we need only exhibit a single GADT whose deep
induction rule cannot be obtained using the techniques of this paper.

@@

By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
although that (necessarily unique) type cannot always be inferred.
The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
data type to represent failure of type inference. It is defined
by:\looseness=-1

@@

This paper extends deep induction to deep GADTs that are not truly
nested. It also shows that truly nested GADTs, deep or not, do not
admit (deep) induction rules. Our development is implemented in Agda,
as is a case study showing how deep induction can prove properties of
GADTs that are not provable by structural induction.\looseness=-1

>>>>

\def\lastname{Johann and Ghiorzi}

@@

\begin{abstract} 
  Deep data types are those that are contructed from other data types,
  including, possibly, themselves.  In this case, they are said to be
  truly nested.  Deep induction is an extension of structural
  induction that traverses {\em all} of the structure in a deep data
  type, propagating predicates on its primitive data throughout the
  entire structure.  Deep induction can be used to prove properties of
  nested types, including truly nested types, that cannot be proved
  via structural induction.  In this paper we show how to extend deep
  induction to deep GADTs that are not truly nested GADTs. We also
  show that the techniques developed in this paper do not suffice for
  extending deep induction to truly nested GADTs.\looseness=-1
\end{abstract}

@@

The inductive data types handled by Coq include (possibly mutually
inductive) polynomial algebraic data types (ADTs), and the induction
rules Coq generates for them are the expected ones for standard
structural induction. It has long been understood, however, that these
rules are too weak to be genuinely useful for deep
ADTs~\cite{jp20}.\footnote{An ADT/nested type/GADT is {\em deep} if it
  is (possibly mutually inductively) defined in terms of other
  ADTs/nested types/GADTs (including, possibly, itself).}  The
following data type of rose trees, here coded in Agda and defined in
terms of the standard type $\mathsf{List}$ of lists (see
Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:

@@

\noindent
%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
%element of its argument list, and similarly for the lifting
%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
list (resp., rose tree) argument.\footnote{Predicate liftings such as
  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
  supplied as primitives or generated automatically from their
  associated data type definitions as described in
  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
  a container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$
  simply traverses containers of that type and applies its predicate
  argument pointwise to the constituent data of type $\mathsf{A}$.
  The ability to define predicate liftings for more general data types
  will be critical to deriving their deep induction rules in
  Section~\ref{sec:framework}.}

Deep induction was also shown in~\cite{jp20} to
% be the missing piece making it possible to
%be the theretofore missing component needed to
% be the essential ingredient making it possible to
deliver the first-ever induction rules --- structural or otherwise ---
for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
types. Deep induction for ADTs and (truly) nested types is reviewed in
Section~\ref{sec:ADTs-and-nesteds} below.

This paper shows how to extend deep induction to proper GADTs, i.e.,
GADTs that go beyond nested types (and thus also beyond ADTs).  A
constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
nested type, take as arguments data whose types involve instances of
$\mathsf{G}$ other than the one being defined. These can even include
instances involving $\mathsf{G}$ itself. If $\mathsf{G}$ is a proper
GADT then at least one of its constructors will also have such a
structured instance of $\mathsf{G}$ --- albeit one not involving
$\mathsf{G}$ itself --- as its codomain.
%But while the types of the arguments to a GADT's constructor can
%include instances of $\mathsf{G}$ that themselves involve
%$\mathsf{G}$, its return type cannot.
For example, the constructor $\mathsf{pair}$ for the GADT

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of pairs, rather than sequences of arbitrary
type, as does $\mathsf{const}$. If one or more of the constructors for
a GADT $\mathsf{G}$ return structured instances of $\mathsf{G}$, then
the GADT will have two distinct, but equally natural, semantics: a
functorial semantics interpreting it as a left Kan
extension~\cite{mac71}, and a parametric semantics interpreting it as
the interpretation of its Church encoding~\cite{atk12,vw10}. As
detailed in~\cite{jgj21}, a key difference in the two semantics is
that the former views GADTs as their {\em functorial
  completions}~\cite{jp19}, and thus as containing more data than just
those expressible in syntax. By contrast, the latter views them as
what might be called {\em syntax-only} GADTs. Happily, these two views
of GADTs coincide for those GADTs that are ADTs or other nested
types. However, both they and their attendant properties differ
greatly for proper GADTs. In fact, the functorial and parametric
semantics for proper GADTs are sufficiently disparate that, by
contrast with the semantics customarily given for ADTs and other
nested types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how
to define a functorial parametric semantics for them~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so GADTs viewed as their functorial completions
do not obviously support induction rules.  This makes sense
intuitively: induction is a syntactic proof technique, so it may not
be possible to use it to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs.  Indeed, this
paper gives the first-ever induction rules --- deep or otherwise ---
for proper GADTs. But it actually delivers far more: it gives a
general framework for deriving deep induction rules for deep GADTs
directly from their syntax. This framework can serve as a basis for
extending modern proof assistants' automatic generation of structural
induction rules for ADTs to automatic generation of deep induction
rules for GADTs. In addition, as for ADTs and other nested types, the
structural induction rule for any GADT can be recovered from its deep
induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

Significantly, deep induction rules for GADTs cannot be derived by
somehow extending the approach of~\cite{jp20} to syntax-only
GADTs. Indeed, the approach taken there makes crucial use of the
functoriality of data types' interpretations from~\cite{jp19}, and
functoriality is precisely what interpreting GADTs as the
interpretations of their Church encodings fails to deliver. Our
approach is to instead first give a predicate lifting styled after
those of~\cite{jp20}, together with a (deep) induction rule, for the
simplest --- and arguably most important --- GADT, namely the equality
GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
the equality GADT to represent $\mathsf{G}$ as its so-called {\em
  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
  ii}) using the predicate liftings for the equality GADT and the
other GADTs appearing in the definition of $\mathsf{G}$ to
appropriately thread the custom predicates for the primitive types
appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
two-step process delivers deep induction rules for a very general
class of deep GADTs. In Section~\ref{sec:GADTs} we introduce a series
of increasingly complex GADTs as running examples, and in
Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
each of them. In particular, we derive the deep induction rule for the
$\mathsf{Seq}$ data type in~\eqref{eq:seq} in
Section~\ref{sec:ind-seq}. We present our general framework for
deriving (deep) induction rules for (deep) GADTs in
Section~\ref{sec:framework}, and observe that the derivations in
Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
Section~\ref{sec:GADT-nested} we show that, by contrast with truly
nested types, which do have a functorial semantics, syntax-only GADTs'
lack of functoriality means that it is not clear how to extend
induction --- deep or otherwise --- to {\em truly nested GADTs}, i.e.,
to proper GADTs whose recursive occurrences appear below
themselves.\footnote{Note carefully the distinction between a GADT
  that is not a nested type --- i.e., a proper GADT --- and a proper
  GADT that is not a truly nested GADT.  In fact, truly nested types
  are not proper GADTs and truly nested GADTs are not (truly) nested
  types. Confusion is understandable in light of the original, and now
  well-established, use of the term ``nested type'' to refer to types
  that do not necessarily involve actual nesting~\cite{bm98}.}  This
does not appear to be much of a restriction, however, since truly
nested GADTs do not, to our knowledge, appear in practice or in the
literature. Section~\ref{sec:app} comprises a case study in using deep
induction. All of the deep induction rules appearing in this paper
have been derived by instantiating our general framework. Our Agda
code implementing them is available at~\cite{web-page}.

@@

{\bf Additional Related Work\/} Various techniques for deriving
induction rules for data types that go beyond ADTs have been
studied. For example, Fu and Selinger~\cite{fs18} show, via examples,
how to derive induction rules for arbitrary nested
types. Unfortunately, however, their technique is rather {\em ad hoc},
so is unclear how to generalize it to nested types other than the
specific ones studied there. Moreover,~\cite{fs18} actually derives
induction rules for data types {\em related} to the original nested
types rather than for the original nested types themselves, and it is
unclear whether or not the derived rules are sufficiently expressive
to prove all results about the original nested types that we would
expect to be provable by induction. This latter point echoes the issue
with Coq-derived induction rule for rose trees raised in
Section~\ref{sec:intro}, which has the unfortunate effect of forcing
users to manually write induction (and other) rules for such types for
use in that system. Tassi~\cite{tas19} derives induction rules for
data type definitions in Coq using unary parametricity. His technique
seems to be essentially equivalent to that of~\cite{jp19} for nested
types, although he does not permit true nesting. More recently,
Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
induction rules for nested types. This plugin is also based on unary
parametricity, and true nesting still is not permitted.  As far as we
know, no attempts have been made to extend either implementation to
truly nested types or to proper GADTs or their deep variants. In fact,
we know of no work other than that reported here that specifically
addresses induction rules for such data types.

@@

Using deep induction we can extend structural induction to {\em truly
  nested types}, i.e., to nested types whose recursive occurrences
appear below themselves. The quintessential example of such a type is
that of bushes\footnote{To define truly nested types in Agda we must
  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
  functions over them we must use the $\mathsf{TERMINATING}$ flag.
  (Similar flags are required in Coq.) Although as programmers we know
  from the metatheory in~\cite{jp19} that $\mathsf{Bush}$ is
  well-defined and the functions we define over them terminate, the
  flags are necessary because Agda fails to infer these
  facts. Analogous comments apply at several places
  below.}\cite{bm98}:

@@

\noindent
Even defining a structural induction rule for bushes requires that we
be able to lift the rule's polymorphic predicate argument to
$\mathsf{Bush}$ itself. This observation was, in fact, the original
motivation for the development of deep induction in~\cite{jp20}. The
deep induction rule for bushes is

@@

We note that the structural induction rule for any (truly) nested type
can be obtained as the special case of its deep induction rule in
which the custom predicates are taken to be constantly
$\mathsf{\top}$-valued predicates. This instantiation ensures that
the resulting induction rule only inspects the top-level structure of
its argument, rather than the contents of that structure, which is
exactly what structural induction should do.
%A concrete example of such a derivation is given in
%Section~\ref{sec:ind-equal}.

Under some circumstances deep induction can be mimicked by
hand-threading applications of structural induction through the layers
of data comprising a deep data type. But this is not the case if,
e.g., one or more of the custom predicates in the data type's deep
induction rule cannot be defined by structural induction.

@@

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, each of $\mathsf{pleaf}$ and
$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
regardless of the instances of $\mathsf{PTree}$ appearing in the types
of its arguments. GADTs relax this restriction, allowing their data
constructors both to take as arguments \emph{and return as results}
data whose types involve instances other than the one being
defined. That is, GADTs' constructors' return type instances can, like
that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.  For every
GADT in this paper, we require that the instance of the return type
for each of its data constructors is a polynomial in that
constructor's type arguments.\looseness=-1
%these instances can be structured.
%
%Applications of GADTs include generic programming, modeling
%programming languages via higher-order abstract syntax, maintaining
%invariants in data structures, and expressing constraints in embedded
%domain-specific languages. GADTs have also been used, e.g., to
%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

@@

Although $\mathsf{Seq}$ does not at first glance appear to be a deep
GADT, when written as its Henry Ford encoding, it, like all GADTs, can
be regarded as ``deep over $\mathsf{Equal}$''.  By contrast, the
%Neither $\mathsf{Equal}$ nor $\mathsf{Seq}$ is a deep GADT, but the
 following GADT $\mathsf{LTerm}$, inspired by~\cite{cis194}, is {\em
   inherently deep}. It encodes terms of a simply typed lambda
 calculus. More robust variations on $\mathsf{LTerm}$ are, of course,
 possible, but this variation is rich enough to illustrate all
 essential aspects of deep GADTs --- and later, in
 Section~\ref{sec:ind-lam}, their deep induction rules --- while still
 being small enough to ensure clarity of exposition.

@@

The equality constraints engendered by GADTs' data constructors makes
deriving (deep) induction rules for them more involved than for ADTs
and other nested types. Nevertheless, we show in this section how to
do so. We first illustrate the key components of our approach by
deriving deep induction rules for the three specific GADTs introduced
in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
abstract these to a general framework that can be applied to any deep
GADT that is not a truly nested GADT. As hinted above, the predicate
lifting for the equality GADT plays a central role in deriving both
structural and deep induction rules for more general GADTs.

@@

We can generalize the approach taken in
Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
deep induction rules for deep GADTs that are not truly nested
GADTs. We will treat deep GADTs of the form

@@

\noindent
for $\mathsf{H^{\wedge}}$. A concrete way to define $\mathsf{H^\wedge
  Map}$ is by induction on the structure of the type $\mathsf{H}$, but
we omit such details since they are not essential to the present
discussion. A further requirement that applies to all of the type
constructors appearing in the right-hand side of the above grammar,
including those in $\ol{\mathsf{K}}$, is that they must all admit
predicate liftings. This is not an overly restrictive condition,
though: all
%types constructed from sums, products, arrow types and
%type application admit predicate liftings, and so do
GADTs constructed from the above grammar admit predicate liftings.
(In particular, the lifting for each type constructor $\mathsf{H}$ is
constructed using its map function.)
%In fact, we have seen such liftings for
%products and type application in Section~\ref{sec:deep-ind-GADTs}.
A concrete way to define more general predicate liftings is, again, by
induction on the structure of the types in a suitable calculus. This
will ensure that the liftings satisfy the crucial property needed to
derive deep induction rules, namely that of distributing over the type
constructors. We do not give a general definition of predicate
liftings here, though, since that would require us to first design a
full type calculus, which is beyond the scope of the present paper.

@@

%{\em Truly nested GADTs} are proper GADTs whose recursive occurrences
%appear below themselves.
In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
derived deep induction rules for GADTs that are not truly nested
GADTs. Since both nested types and GADTs without true nesting admit
deep induction rules, we might expect truly nested GADTs to admit them
as well. Unfortunately, however, the techniques developed in the
previous sections do not extend to truly nested GADTs. Indeed, the
induction rule for a data type generally relies on (unary)
parametricity of the model interpreting it, and deep induction for a
truly nested type or a truly nested GADT also relies on this
interpretation being functorial.  Whereas ADTs and nested types both
admit functorial parametric semantics, proper GADTs admit parametric
semantics but do not admit functorial semantics. In this section we
show how the techniques developed in this paper for deriving deep
induction rules go wrong for truly nested GADTs by analyzing the
following very simple example:\looseness=-1

@@

\begin{equation*}\label{gadt-nested}
\begin{array}{l}
\mathsf{data\ G : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;\;c :\,  \forall \{A : Set\} \to G\,(G\,A) \to G\,(A \times A)}
\end{array}
\end{equation*}
We acknowledge that $\mathsf{G}$ is semantically equivalent to the
empty data type, and thus has a trivial (deep) induction principle. We
could, of course, consider a more realistic counterexample, but this
would only add notational overhead for no gain in conceptual
clarity. Indeed, we need only exhibit a single GADT whose deep
induction rule cannot be obtained using the techniques of this paper,
and for which more robust techniques will therefore be needed if deep
induction rules are to be derived for them.

@@

By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
but that (necessarily unique) type cannot always be inferred.  The
predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$ data
type to represent failure of type inference. It is defined
by:\looseness=-1

@@

This paper extends deep induction to deep GADTs that are not truly
nested. It also shows that truly nested GADTs, deep or not, do not
obviously admit (deep) induction rules. Our development is implemented
in Agda, as is our case study from Section~\ref{sec:app}.
